Vagueness and imprecise imitation in signaling games

### 2016-02-27 paper/version_05/paper.tex
sections 1-2 reshuffled; partial pooling paragraph added

But there are other intriguing aspects about vagueness. One perplexing issue is how vagueness
could arise and be maintained in the first place. This is an apparent puzzle for functionalist
accounts that maintain that concepts and linguistic meanings evolved driven towards
efficiency. \citet{Lipman2009:Why-is-Language} argues that, technically, the existence of unclear borderline cases always entails inefficiency of
categorization or communication. The challenge is then to explain how
vagueness can persist both: (i) under evolutionary pressure to be optimally discriminative, and (ii) without
jeopardizing the possibility of evolving, learning and communicating with a meaningful
language. A number of authors have consequently tried to explain why vagueness evolved as
something that is itself useful
\citep[e.g.][]{Jaegherde-Jaegher2003:A-Game-Theoreti,Deemter2009:Utility-and-Lan,BlumeBoard2013:Intentional-Vag}.
Others have argued that vagueness is a natural byproduct of limitations in information
processing \citep[e.g.][]{FrankeJager2010:Vagueness-Signa} or of generalization in low-level
learning strategies \citep[e.g.][]{OConnor2013:The-Evolution-o}. This paper contributes to the
latter line of thought.

We believe that vagueness in language and thought may have many reasons, not just one. We focus
here on one \emph{a priori} plausible reason for why vagueness is natural and pervasive. The
idea is that at least some meanings are vague, in part because of imprecision in the perception
of similar stimuli and in the realization of similar responses.\footnote{On this view,
  vagueness in language may be seen as a necessary sub-optimality due to limitations in another
  domain of cognition, first and foremost perception. The whole system, e.g., perception and
  language together, may be an almost optimal adaptation in a larger frame of reference, e.g.,
  if we take into account the metabolic costs for increased accuracy.} While this idea may be
\emph{prima facie} plausible, it is not immediately clear what the consequences of
confusability of stimuli are on a process of meaning evolution. In other words, in order to
address Lipman's challenge seriously, and not just hand-wave it away, it needs to be spelled
out how exactly confusability of stimuli could come into play in a process of meaning evolution
and how it could lead to vague but by-and-large informative signal meaning. The formal model
that this paper introduces does exactly that. And more. % We find that confusability of stimuli
% can regularize and systematize evolving meaning, and may thereby imply an advantageous
% side-effect that may counter some of the disadvantages that imprecise and vague communication
% may have. Here is some elaboration on this point.

The emphasis of previous models that studied the effects of transmission infidelity on the
evolution of language has been on inductive biases and the systematicity, compression and
regularization that they can introduce. Here, we would like to show that transmission noise of
a different kind can lead to regularization as well and also give rise to vague
meaning. Towards this end, we formalize the expected change in the behavior of a population of
agents that try to imitate other agents' signaling behavior. We assume, however, that both
realization of behavior, as well as observability of others' behavior, is molested by noise, in
particular the probabilistic confusability of similar stimuli. The resulting population-level
dynamic generalizes the replicator dynamic \citep{TaylorJonker1978:Evolutionary-St} in its
interpretation as a cultural evolutionary dynamic based on imitation
\citep[e.g.][]{Helbing1996:A-Stochastic-Be,Schlag1998:Why-Imitate-and}. The inclusion of
confusability of stimuli does not undermine the possibility of evolving communicative signaling
behavior. Instead, it leads to the evolution of vague meanings. It moreover accelerates the
emergence of communicative signaling because it unifies and regularizes evolutionary outcomes,
making it appear as if agents were applying inductive biases or generalizing over partial
observations, when in truth this is the sole effect of confusion of similar stimuli.

@@

The next section introduces the background against which the work presented here can be
appreciated. Section~\ref{sec:repl-diff-dynam} introduces our generalization of the replicator
dynamic which is derived from the idea that agents imitate other agents' behavior while
possibly confusing similar states. This imprecise imitation dynamic is explored in
Section~\ref{sec:exploring-rdd}. Section~\ref{sec:discussion} reflects and compares our
approach to others. Appendix~\ref{sec:formal-stuff} provides formal detail.

\mf{intro misses a mentioning of signaling games}

@@

@@

>>>>

But there are other intriguing aspects about vagueness. One perplexing issue is how vagueness
could arise and be maintained in the first place. This is an apparent puzzle for functionalist
accounts that maintain that concepts and linguistic meanings evolved driven towards
efficiency. \citet{Lipman2009:Why-is-Language} argues that, technically, the existence of unclear borderline cases always entails inefficiency of
categorization or communication. The challenge is then to explain how
vagueness can persist both: (i) under evolutionary pressure to be optimally discriminative, and (ii) without
undermining the possibility of evolving, learning and communicating with a meaningful
language. A number of authors have consequently tried to explain why vagueness evolved as
something that is itself useful
\citep[e.g.][]{Jaegherde-Jaegher2003:A-Game-Theoreti,Deemter2009:Utility-and-Lan,BlumeBoard2013:Intentional-Vag}.
Others have argued that vagueness is a natural byproduct of limitations in information
processing \citep[e.g.][]{FrankeJager2010:Vagueness-Signa} or of generalization in low-level
learning strategies \citep[e.g.][]{OConnor2013:The-Evolution-o}. This paper contributes to the
latter line of thought.

We believe that vagueness in language and thought may have many reasons, not just one. We focus
here on one \emph{a priori} plausible reason for why vagueness is natural and pervasive. The
idea is that vagueness is, at least in part, due to imprecision in the perception of similar
stimuli and imprecision in the realization of similar responses. On this view, vagueness in
language may be seen as a necessary sub-optimality due to limitations in another domain of
cognition, first and foremost perception. It could then be speculated that the whole system,
e.g., perception and language together, may be an almost optimal adaptation in a larger frame
of reference, e.g., if we take into account the metabolic costs for increased perceptual
accuracy. We will not engage in such speculation here. Rather we will explore the consequences
of perceptual limitations on processes of meaning evolution in a suitable formal framework. In
other words, in order to address Lipman's challenge seriously, and not just hand-wave it away
by appeal to the naturalness of vagueness, it needs to be spelled out how exactly confusability
of stimuli could come into play in a process of meaning evolution and how it could lead to
vague but by-and-large informative signal meaning. The formal model that this paper introduces
does exactly that. But it also does more. We find that confusability of stimuli can regularize
and systematize evolving meaning. This suggests a possibly advantageous side-effect that a
natural cause of vagueness may have and that might counter some of the disadvantages that
vagueness may have for communicative efficiency \citep[c.f.][]{OConnor2013:The-Evolution-o}.

@@

The next section introduces the background against which the work presented here can be
appreciated. Section~\ref{sec:repl-diff-dynam} introduces a generalization of the replicator
dynamic which is derived from the idea that agents imitate other agents' behavior while
possibly confusing similar states. This imprecise imitation dynamic is explored in
Section~\ref{sec:exploring-rdd}. Section~\ref{sec:discussion} reflects and compares our
approach to others. Appendix~\ref{sec:formal-stuff} provides formal detail.

@@

To illustrate, let us briefly consider a sim-max game with six equiprobable states and two
messages, its equilibria and what vagueness intuitively would look like \citep[see][for a more
thorough discussion]{Jager2007:The-Evolution-o,OConnor2013:The-Evolution-o}. We assume that
utilities are strictly decreasing with decreasing
similarity. Figure~\ref{fig:EquilibriaVagueness} shows three pairs of sender and receiver
strategies. States and acts are arranged according to their similarity: the closer they are to
each other, the more similar they are. The pair in Figure~\ref{fig:ineffNonEq} is not an
equilibrium, because the sender's non-convex use of signals is suboptimal given the receiver's
behavior. In contrast, Figure~\ref{fig:efficientEq}Z shows a maximally efficient
equilibrium. This is a partial pooling equilibrium in the sense that the speaker uses the same
message for several states. This would be inefficient if there were more messages, but it is
maximally efficient for the two-message case. So, while partial pooling may entail
inefficiency, and while partial pooling can hamper the evolution of maximally efficient signal
use
\citep[e.g.][]{Hutteger:2007_Evol_Indicatives_Imperatives,Pawlowitsch2008:Why-Evolution-d,HutteggerSkyrms2010:Evolutionary-Dy},
this is orthogonal to our concerns about vagueness. A regular and natural yet vague signal use
would look like in Figure~\ref{fig:vagueUse}. This is not an equilibrium, but it gets close, so
to speak. It shows smooth transitions across similar states at the boundaries of categories and
across acts around the most prototypical instances (as indicated by decreasing thickness of
arrows in Figure~\ref{fig:vagueUse}).

  \begin{subfigure}[]{0.475\textwidth}
    \centering
    \begin{tikzpicture}

      \draw[] (0,2) node (t1) {$\mystate{1}$};

      \draw[] (1,2) node (t2) {$\mystate{2}$};

      \draw[] (2,2) node (t3) {$\mystate{3}$};

      \draw[] (3,2) node (t4) {$\mystate{4}$};

      \draw[] (4,2) node (t5) {$\mystate{5}$};

      \draw[] (5,2) node (t6) {$\mystate{6}$};

      \draw[] (1,1) node (m1) {$\mymessg{1}$};

      \draw[] (4,1) node (m2) {$\mymessg{2}$};

      \draw[] (0,0) node (a1) {$\myact{1}$};

      \draw[] (1,0) node (a2) {$\myact{2}$};

      \draw[] (2,0) node (a3) {$\myact{3}$};

      \draw[] (3,0) node (a4) {$\myact{4}$};

      \draw[] (4,0) node (a5) {$\myact{5}$};

      \draw[] (5,0) node (a6) {$\myact{6}$};

      % arrows

      \draw[->, thick] (t1) -- (m1);

      \draw[->, thick] (t2) -- (m1);

      \draw[->, thick] (t3) -- (m2);

      \draw[->, thick] (t4) -- (m1);

      \draw[->, thick] (t5) -- (m2);

      \draw[->, thick] (t6) -- (m2);

      \draw[->, thick] (m1) -- (a2);

      \draw[->, thick] (m2) -- (a5);

    \caption{inefficient, non-convex, non-equilibrium}
    \label{fig:ineffNonEq}
  \end{subfigure}
  \hfill
  \begin{subfigure}[]{0.475\textwidth}
    \centering

    \begin{tikzpicture}

      \draw[] (0,2) node (t1) {$\mystate{1}$};

      \draw[] (1,2) node (t2) {$\mystate{2}$};

      \draw[] (2,2) node (t3) {$\mystate{3}$};

      \draw[] (3,2) node (t4) {$\mystate{4}$};

      \draw[] (4,2) node (t5) {$\mystate{5}$};

      \draw[] (5,2) node (t6) {$\mystate{6}$};

      \draw[] (1,1) node (m1) {$\mymessg{1}$};

      \draw[] (4,1) node (m2) {$\mymessg{2}$};

      \draw[] (0,0) node (a1) {$\myact{1}$};

      \draw[] (1,0) node (a2) {$\myact{2}$};

      \draw[] (2,0) node (a3) {$\myact{3}$};

      \draw[] (3,0) node (a4) {$\myact{4}$};

      \draw[] (4,0) node (a5) {$\myact{5}$};

      \draw[] (5,0) node (a6) {$\myact{6}$};

      % arrows

      \draw[->, thick] (t1) -- (m1);

      \draw[->, thick] (t2) -- (m1);

      \draw[->, thick] (t3) -- (m1);

      \draw[->, thick] (t4) -- (m2);

      \draw[->, thick] (t5) -- (m2);

      \draw[->, thick] (t6) -- (m2);

      \draw[->, thick] (m1) -- (a2);

      \draw[->, thick] (m2) -- (a5);

    \caption{maximally efficient equilibrium, partial pooling}
    \label{fig:efficientEq}
  \end{subfigure}

\begin{subfigure}[]{0.475\textwidth}
    \centering

    \begin{tikzpicture}

      \draw[] (0,2) node (t1) {$\mystate{1}$};

      \draw[] (1,2) node (t2) {$\mystate{2}$};

      \draw[] (2,2) node (t3) {$\mystate{3}$};

      \draw[] (3,2) node (t4) {$\mystate{4}$};

      \draw[] (4,2) node (t5) {$\mystate{5}$};

      \draw[] (5,2) node (t6) {$\mystate{6}$};

      \draw[] (1,1) node (m1) {$\mymessg{1}$};

      \draw[] (4,1) node (m2) {$\mymessg{2}$};

      \draw[] (0,0) node (a1) {$\myact{1}$};

      \draw[] (1,0) node (a2) {$\myact{2}$};

      \draw[] (2,0) node (a3) {$\myact{3}$};

      \draw[] (3,0) node (a4) {$\myact{4}$};

      \draw[] (4,0) node (a5) {$\myact{5}$};

      \draw[] (5,0) node (a6) {$\myact{6}$};

      % arrows

      \draw[->, thick] (t1) -- (m1);

      \draw[->, thick] (t2) -- (m1);

      \draw[->, thick, dashed] (t3) -- (m1);

      \draw[->, thick, dotted] (t3) -- (m2);

      \draw[->, thick, dotted] (t4) -- (m1);

      \draw[->, thick, dashed] (t4) -- (m2);

      \draw[->, thick] (t5) -- (m2);

      \draw[->, thick] (t6) -- (m2);

      \draw[->, thick, dotted] (m1) -- (a1);

      \draw[->, thick, dashed] (m1) -- (a2);

      \draw[->, thick, dotted] (m1) -- (a3);

      \draw[->, dotted] (m1) -- (a4);

      \draw[->, dotted] (m2) -- (a3);

      \draw[->, thick, dotted] (m2) -- (a4);

      \draw[->, thick, dashed] (m2) -- (a5);

      \draw[->, thick, dotted] (m2) -- (a6);

    \caption{almost efficient, probabilistic, intuitively vague}
    \label{fig:vagueUse}
  \end{subfigure}

  \caption{Examples of strategy pairs in sim-max games.}
  \label{fig:EquilibriaVagueness}
\end{figure}

@@

The emphasis of previous models that studied the effects of transmission infidelity on the
evolution of language has been on inductive biases and the systematicity, compression and
regularization that they can introduce. Here, we would like to show that transmission noise of
a different kind can lead to regularization as well and also give rise to vague
meaning. Towards this end, we formalize the expected change in the behavior of a population of
agents that try to imitate other agents' signaling behavior. We assume, however, that both
realization of behavior, as well as observability of others' behavior, is systematically
perturbed by noise. The resulting population-level dynamic generalizes the replicator dynamic
\citep{TaylorJonker1978:Evolutionary-St} in its interpretation as a cultural evolutionary
dynamic based on imitation
\citep[e.g.][]{Helbing1996:A-Stochastic-Be,Schlag1998:Why-Imitate-and}. The inclusion of
confusability of stimuli does not undermine the possibility of evolving communicative signaling
behavior. Instead, it leads to the evolution of vague meanings. It moreover accelerates the
emergence of communicative signaling because it unifies and regularizes evolutionary outcomes,
making it appear as if agents were applying inductive biases or generalizing over partial
observations, when in truth this is the sole effect of confusion of similar stimuli.
